Staged Abstract Interpreters

### 2019-04-05 paper/10_intro.tex
paper

We show that these well-known abstractions can be combined smoothly, and more
importantly almost free. Through staging, the major overhead of
abstractions -- from the monadic layers -- is completely eliminated in the
generated code. \todo{put some numbers} However, we do not lose any advantages
of this high-level programming style (e.g., modularity and extensibility); the
meaning of the program also remains the same, particularly, the soundness of 
static analysis after staging.

\paragraph{Performance via Specialization}

Besides the intellectual merit shown in the confluence diagram, this paper
contributes to a practical problem in static analysis: \textit{can we speed up a
static analyzer without any intrusive changes and soundness compromises?} Static
analysis is known as a trade-off game between \textit{precision} and
\textit{efficiency}, we argue that applying meta-programming (multi-stage
programming, in particular) with abstract interpreters is an effective approach
to improve the latter while the former is untouched.
In the literature, specializing analysis with respect to the input program
\cite{damian1999partial, amtoft1999partial, Boucher:1996:ACN:647473.727587,
ashley:practical} is not new, despite the fact that most of the previous work
are ad-hoc on a particular analysis or requiring significant changes to the
analyzer. For example, abstract compilation is one of the techniques
\cite{Boucher:1996:ACN:647473.727587}. But it requires the whole analyzer to be
rewritten to the closure generation form. In this paper, we advocate a systematic
approach to meta-programming that achieves the performance goal with less
intrusive changes; in fact, a case study (Section~\ref{cs_ac}) comparing our
approach with abstract compilation shows that, by utilizing type-based stage
annotations, the analyzer program does not need any changes.

\paragraph{Modularity via Meta-Programming} \todo{compress this part}

Our next contribution is to enable modular analysis based on a whole-program
analyzer by meta-programming. Modern software are shipped with large libraries
\cite{toman_et_al:LIPIcs:2017:7121}, and analyzing programs with large libraries
is usually expensive or unnecessarily repeated for whole-program analyzing
algorithms. For example, it has been shown that analyzing a simple ``Hello
World'' program in Java depends on additional 3,000 classes in the library
\cite{DBLP:conf/oopsla/KulkarniMZN16}. A natural idea is to analyze libraries
and generate summaries, later we can plugin the necessary information into the
summaries and finish the analysis; therefore such summaries should be modular
and reusable. Meta-programming gives us a chance to readily tweak a
whole-program analyzing algorithm into an analyzer that works modularly and
produces summaries. Still, the idea is to specialize the abstract interpreter
that was used for whole-program analysis, but on a subset of the program, for
examples, a set of generic functions for list manipulation. Such specialization
works on lambda expressions and generates code, which is partial analysis
results and can be completed by running with the necessary arguments (for
example, an environment that closes the lambda term) at next stages. Multiple
partial analysis results also can be reused later and composed together as they
are just normal programs at the next stage. Through staging, we mechanically
obtain a modular analysis, even though the original algorithm is formulated as a
whole-program analysis.

>>>>

We show that these well-known abstractions can be combined smoothly, and more
importantly almost free. Through staging, the major overhead of
abstractions -- from the monadic layers -- is completely eliminated in the
generated code. We observe at most 14x times faster than the unstage abstract
interpreter.
However, we do not lose any advantages of this high-level programming style
(e.g., modularity and extensibility); the meaning of the program also remains
the same, particularly, the soundness of static analysis after staging.

%\paragraph{Performance via Specialization}
Besides the intellectual merit shown in the confluence diagram, this paper
contributes to a practical problem in static analysis: \textit{can we speed up a
static analyzer without any intrusive changes and soundness compromises?} Static
analysis is known as a trade-off game between \textit{precision} and
\textit{efficiency}, we argue that applying meta-programming (multi-stage
programming, in particular) with abstract interpreters is an effective approach
to improve the latter while the former is untouched.
In the literature, specializing analysis with respect to the input program
\cite{damian1999partial, amtoft1999partial, Boucher:1996:ACN:647473.727587,
ashley:practical} is not new, despite the fact that most of the previous work
are ad-hoc on a particular analysis or requiring significant changes to the
analyzer. For example, abstract compilation is one of the techniques
\cite{Boucher:1996:ACN:647473.727587}. But it requires the whole analyzer to be
rewritten to the closure generation form. In this paper, we advocate a systematic
approach of meta-programming that achieves the performance goal with less
intrusive changes and engineering efforts.

\paragraph{Applications}

Further more, we show various applications of staging an abstract interpreter.
(1) We compare our approach with abstract compilation, and show that we can
achieve the same optimization by utilizing type-based stage annotations,
meanwhile the analyzer program does not need any changes. (2) \todo{} (3)
Staging abstract interpreter enables modular analysis and provides a mechanized
way to create library summaries. We notice that we may apply the staged abstract
interpreter separately on the different part of the analyzed program, the
generated code is reusable and composable. Therefore, we can turn a
whole-program analyzer into a analyzer that runs modularly.

\iffalse
Our next contribution is to enable modular analysis based on a whole-program
analyzer by meta-programming. Modern software are shipped with large libraries
\cite{toman_et_al:LIPIcs:2017:7121}, and analyzing programs with large libraries
is usually expensive or unnecessarily repeated for whole-program analyzing
algorithms. For example, it has been shown that analyzing a simple ``Hello
World'' program in Java depends on additional 3,000 classes in the library
\cite{DBLP:conf/oopsla/KulkarniMZN16}. A natural idea is to analyze libraries
and generate summaries, later we can plugin the necessary information into the
summaries and finish the analysis; therefore such summaries should be modular
and reusable. Meta-programming gives us a chance to readily tweak a
whole-program analyzing algorithm into an analyzer that works modularly and
produces summaries. Still, the idea is to specialize the abstract interpreter
that was used for whole-program analysis, but on a subset of the program, for
examples, a set of generic functions for list manipulation. Such specialization
works on lambda expressions and generates code, which is partial analysis
results and can be completed by running with the necessary arguments (for
example, an environment that closes the lambda term) at next stages. Multiple
partial analysis results also can be reused later and composed together as they
are just normal programs at the next stage. Through staging, we mechanically
obtain a modular analysis, even though the original algorithm is formulated as a
whole-program analysis.
\fi

### 2019-04-05 paper/40_absinterp.tex
paper

In this section, we simply let @R[T] = T@ (i.e., @NoRep@) to instantiate the
unstaged lattice type class. An simple instance of lattice we used is the power
set abstract domain (shown in Figure \ref{fig:powerset}), where the ordering is subset relation, the join is union
operation, and the meet is intersect operation.
Other lattices used in the remainder of the paper, such as products and maps, can be
implemented similarly, or lifted from existing lattices element-wise or
point-wise.

@@

The sketch implementation of @SetT@ transformer is shown below. The value @run@
it encapsulates is of type @M[SetT[A]]@, where @M[_]@ is the inner monad, and
@A@ the element type of the list. The @flatMap@ method first unwraps the inner
monadic value @M[SetT[A]]@ and obtain a value of @Set[A]@, then applies @f@ to
each element in that list, and finally folds the results back to a single
@SetT[M, B]@.

>>>>

In this section, we simply let @R[T] = T@ (i.e., @NoRep@) to instantiate the
unstaged lattice type class. An simple instance of lattice we used is the power
set abstract domain (shown in Figure \ref{fig:powerset}), where the ordering is subset relation, the join is union
operation, and the meet is intersect operation.
Other lattices used in the remainder of the paper, such as products and maps, can be
implemented similarly, or lifted from existing lattices element-wise or
point-wise. Naturally, abstract domains such intervals also can be implemented
in a stage polymorphic way.

@@

The sketch implementation of @SetT@ transformer is shown below. The value @run@
encapsulated by the monad is of type @M[SetT[A]]@, where @M[_]@ is the inner monad, and
@A@ is the element type of the list. The @flatMap@ method first unwraps the inner
monadic value @M[SetT[A]]@ and obtain a value of @Set[A]@, then applies @f@ to
each element in that list, and finally folds the results back to a single
@SetT[M, B]@.

### 2019-04-05 paper/60_cases.tex
paper

@@

\subsubsection{Abstract Garbage Collection at Staging Time}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%5

\subsubsection{A More Precise Store Model}

The store-widened analysis we implemented in Section~\ref{sai} uses a single
store to approximate the runtime store. It can be efficently computed in
polynomial time together with 0-CFA-like abstraction, but sometimes we may
desire a more precise result that distinguishes the final values and stores for
different closure targets. To achieve this goal, we need to tweak our abstract
interpreter and type instantiation. The answer type @Ans@ is changed to a set of
@VS@s where a @VS@ is a pair of sets of abstract values (such as closures or
abstract numbers) and a store.

\begin{lstlisting}
type VS = (Set[AbsValue], Store)
type Ans = R[Set[VS]]
\end{lstlisting}

Note that the type @Ans@ uses our stage polymorphic type @R@, meaning that under
staging the type @Ans@ represents a next stage value. Then, our generic
interpreter is also changed when handling function applications.

\begin{lstlisting}
case App(e1, e2) =>
  val e1ans = ev(e1, $\rho$, $\sigma$)
  val e1vs  = choices(e1ans)
  val e2ans = ev(e2, $\rho$, e1vs.$\sigma$)
  val e2vs  = choices(e2ans)
  apply_closure(ev)(e1vs.v, e2vs.v, e2vs.$\sigma$)
\end{lstlisting}

The idea to handle the application is to explore all possible closures from @e1@
and meanwhile use the latest store. What @choices@ does is similar to McCarthy's
@amb@ operator ~\cite{MCCARTHY196333}: it non-deterministically returns an
element of type @VS@ from its argument, e.g., @e1ans@, to its right-hand side
receiver, i.e., @e1vs@. The function @choices@ internally uses the delimited
control operator @shift@ to capture the continuation, which is the code block
after its call site. This allows us to perform nesting depth-first evaluation
for function application while still writing the program in direct-style
\cite{Wei:2018:RAA:3243631.3236800}. Again, we can stage this part as we did for
the naive abstract interpreter.

\subsubsection{Context-Sensitivity}

We add k-CFA-like context-sensitivity to the analysis by introducing an abstract
timestamp, whose concrete instantiation is a finite list of expressions that
track $k$ recent calling contexts. The definition of abstract addresses is
changed to a tuple of identifiers and the time it get allocated, meaning that
this address points to some values under such calling context. If $k$ is 0, we
obtain a monovariant analysis as demonstrated before; if $k > 0$, we obtain a
family of analysis with increasing precision.

\begin{lstlisting}
type Time = List[Expr]
\end{lstlisting}

Every time when we call the @eval@ function, we refresh the timestamp by calling
function @tick@, which returns a new timestamp. Here we adopt a $k$-CFA-like
allocation strategy, therefore the @tick@ function can be implemented as
appending the current expression being evaluated to the existing calling
context, and then taking the first $k$ elements from the list.

\begin{lstlisting}
def tick(e: Expr, $\tau$: R[Time]): R[Time] = (e :: $\tau$).takewhile
def eval(ev: EvalFun)
  (e: Expr, $\rho$: R[Env], $\sigma$: R[Store], $\tau$: R[Time]): Ans = {
    val $\tau$_* = tick(e, $\tau$)
    e match {
      case Lit(i) => ...
      ...
    }
  }
\end{lstlisting}

Accordingly, the type of return value is accompanied by the timestamp. For a
recursive call of @ev@, which would also return the latest timestamp, and that
timestamp will be used for the evaluation afterward.

\begin{lstlisting}
type VST = (Value, Store, Time)
type Ans = R[Set[VST]]
\end{lstlisting}

Using other allocation strategies to achieve different sensitivities is also
possible \cite{DBLP:conf/icfp/Gilray0M16} and can be staged under our framework.

@@

\iffalse
Another perspective: programs are data for an abstract interpreter, so if we
have $n$ programs, then maybe there can be $n$ stages. Probably we can analyze
first $m$ programs, and generate a residual abstract interpreter waiting for the
rest $(n-m)$ programs. These $(n-m)$ programs might be (abstract) arguments for
the first $n$ programs, and the abstract interpreter itself might be a partial
abstract interpreter.
\fi

>>>>

The abstract addresses we introduced in Section \ref{sai} is simply identical to
the variable names, which is a context-insensitive schema.
Through tweaking the address allocation strategy, we can achieve various
context-sensitive analysis \cite{DBLP:conf/icfp/Gilray0M16}. In this section, we
demonstrate a calling-context-sensitive analysis, i.e., $k$-CFA-like analysis
\cite{DBLP:journals/jfp/HornM12}, in our staged abstract interpreter.

\begin{lstlisting}
  type Time = List[Expr]
  case class KCFAAddr(x: Ident, time: Time) extends Addr
\end{lstlisting}

The timestamp is essentially a finite list of expressions that tracks the $k$
recent calling contexts. The definition of abstract addresses @KCFAAddr@ is also
changed to contain a identifier and the time when it get allocated, meaning that
this address points to some values under such calling context. If $k$ is 0, we
obtain a monovariant analysis as demonstrated before; if $k > 0$, we obtain a
family of analysis with increasing precision.

\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepStateT[
                   RepStateT[RepSetReaderStateM[Cache, Cache, ?], Time, ?], //Add Time into a RepStateT
                 Store, ?], Env, T]
  type Result = (Rep[Set[(Value, Store, Time)]], Rep[Cache])
\end{lstlisting}

We then integrate the timestamp into our moand stack by adding a staged
@StateT@ monad. The result type is passingly get updated to contain a set of
tuples, where the time is added.

\begin{lstlisting}
  def tick(e: Expr): AnsM[Unit] = for {
    τ <- get_time; u <- liftM(StateTMonad.put((e::τ).take($k$)))
  } yield u
  def alloc(x: String): AnsM[Addr] = for {
    τ <- get_time
  } yield KCFAAddr(x, τ)
\end{lstlisting}

Every time when we call the @eval@ function, we refresh the timestamp by calling
the @tick@ function, which update the timestamp in the state monad. The @tick@
function is implemented as prepending the current expression being evaluated @e@
to the existing calling context, and then taking the first $k$ elements from the list.
The type @Config@ used in caching is also changed to included the timestamp.

@@

An commonly used technique to improve the running time in control-flow analysis
is store-widening. As shown by \cite{Darais:2015:GTM:2814270.2814308,
DBLP:journals/pacmpl/DaraisLNH17}, store-widening analysis can be easily
implemented in monadic abstract interpreter by swaping the @StateT@ (for store)
and the @SetT@ (for nondeterminism). In this section, we show this optimization
is orthogonal to staging and we just need to swap the same \textit{staged}
version of monads. After the adjustment, we have the following monad stack:

\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepSetT[RepStateT[
                   RepReaderT[RepStateT[IdM, Cache, ?], Cache, ?], Store, ?], ?], Env, T]
\end{lstlisting}

\subsubsection{Abstract Garbage Collection}

@@

\iffalse
Another perspective: programs are data for an abstract interpreter, so if we
have $n$ programs, then maybe there can be $n$ stages. Probably we can analyze
first $m$ programs, and generate a residual abstract interpreter waiting for the
rest $(n-m)$ programs. These $(n-m)$ programs might be (abstract) arguments for
the first $n$ programs, and the abstract interpreter itself might be a partial
abstract interpreter.

### 2019-04-05 paper/70_eval.tex
paper

\begin{figure}[h]
\tiny
\begin{tabular}{@{}lllll@{}}
\toprule
    program             &\#AST & unstaged   & staged     & $\frac{\text{unstaged}}{\text{staged}}$ \\ \midrule
    @rsa@               & 451  & 0.308      & 0.064      & 459.15\%          \\
    @church@            & 120  & 71.811     & 23.990     & 299.34\%          \\
    @fermat@            & 310  & 0.043      & 0.006      & 716.67\%          \\
    @mbrotZ@            & 331  & 1.112      & 0.214      & 519.63\%          \\
    @lattice@           & 609  & 31.874     & 8.816      & 358.69\%          \\
    @kcfa-worst-16@     & 182  & 0.083      & 0.013      & 638.46\%          \\
    @kcfa-worst-32@     & 358  & 0.452      & 0.044      & 1027.27\%         \\
    @kcfa-worst-64@     & 774  & 3.329      & 0.236      & 1410.59\%         \\
    @solovay-strassen@  & 523  & 5.94       & 1.189      & 428.43\%          \\
    \bottomrule
\end{tabular}
\caption{Evaluation result for 0CFA without store-widening.} \label{evaluation_result}
\end{figure}

>>>>

\begin{figure}[h]
\footnotesize
\begin{tabular}{@{}lllll@{}}
\toprule
    program             &\#AST & unstaged   & staged     & $\frac{\text{unstaged}}{\text{staged}}$ \\ \midrule
    @fib@               & 32   & 0.013      & 0.002      & 650.00\%          \\
    @rsa@               & 451  & 0.308      & 0.064      & 459.15\%          \\
    @church@            & 120  & 71.811     & 23.990     & 299.34\%          \\
    @fermat@            & 310  & 0.043      & 0.006      & 716.67\%          \\
    @mbrotZ@            & 331  & 1.112      & 0.214      & 519.63\%          \\
    @lattice@           & 609  & 31.874     & 8.816      & 358.69\%          \\
    @kcfa-worst-16@     & 182  & 0.083      & 0.013      & 638.46\%          \\
    @kcfa-worst-32@     & 358  & 0.452      & 0.044      & 1027.27\%         \\
    @kcfa-worst-64@     & 774  & 3.329      & 0.236      & 1410.59\%         \\
    @solovay-strassen@  & 523  & 5.94       & 1.189      & 428.43\%          \\
    \bottomrule
\end{tabular}
\caption{Evaluation result for 0CFA without store-widening.} \label{evaluation_result}
\end{figure}

### 2019-04-05 paper/80_related.tex
paper

\paragraph{Optimizing Static Analysis Through Specialization}
The underlying idea in this paper is closely related to abstract compilation (AC)
\cite{Boucher:1996:ACN:647473.727587}: removing the interpretation overhead on
traversing the syntax tree by specialization. Specifically, the residual program
of AC can be either textual or closures. However, as we studied in Section
\ref{cs_ac}, with the perspective from high-performance DSLs and monadic
interpreters, we make it more general, extensible and easy.
\citet{Johnson:2013:OAA:2500365.2500604} adapt the idea of closure
generation to optimize small-step abstract interpreter in state-transition
style. The analyzed program is firstly compiled to a IR called "abstract
bytecode", which are actually higher-order functions and will be executed later
on a abstract abstract machine for that IR. \citet{damian1999partial} provides a
formal treatment to abstract compilation and Shiver's CFA, as well as proofs to
establish correctness of the certified specialized analyzer.
\citet{amtoft1999partial} applied partial evaluation for constraint-based
control flow analysis. Splitting an analysis into multiple stages is also
studied for analysis other than for control-flow, though the formulation may
very different. For example, \citet{DBLP:conf/cgo/HardekopfL11} apply staging to
flow-sensitive pointer analysis. The first stage is to analyze the program code
to obtain a sparse representation, and then the second stage conducts the
flow-sensitive analysis based on the first one.
The abstract domains can be also specialized w.r.t. the static program
structure, e.g., decomposing polyhedras \cite{DBLP:conf/popl/SinghPV17,
Singh:2017:PCD:3177123.3158143} to smaller ones such that the abstract
transformers can compute using less time and space.

\paragraph{Abstract Interpreters} Abstract interpretation was proposed as a
semantic-based approach to construct sound static analysis by approximation
\cite{DBLP:conf/popl/CousotC77}. As semantic artifacts, the Abstracting Abstract
Machines (AAM) \cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10} approach
shows that abstract interpreters can be derived systematically from concrete
semantic artifacts.
Using monads to construct abstract interpreters is explored by
\citet{Sergey:2013:MAI:2491956.2491979} and
\citet{DBLP:journals/pacmpl/DaraisLNH17, Darais:2015:GTM:2814270.2814308}.
Similar to monadic abstract interpreters, \citet{Wei:2018:RAA:3243631.3236800}
reconstruct big-step abstract interpreters with delimited continuations;
\citet{Keidel:2018:CSP:3243631.3236767} present a modular arrow-based abstract
interpreter that makes proof of soundness can be constructed compositionally.
\citet{DBLP:conf/cc/CousotC02} proposed an abstract interpretation framework for
modular analysis.

>>>>

\paragraph{Optimizing Static Analysis Through Specialization}
The underlying idea in this paper is closely related to abstract compilation (AC)
\cite{Boucher:1996:ACN:647473.727587}: removing the interpretation overhead on
traversing the syntax tree by specialization. Specifically, the residual program
of AC can be either textual or closures. However, as we studied in Section
\ref{cs_ac}, with the perspective from high-performance DSLs and monadic
interpreters, we make it more general, extensible and easy.
\citet{Johnson:2013:OAA:2500365.2500604} adapt the idea of closure
generation to optimize small-step abstract interpreter in state-transition
style. The analyzed program is firstly compiled to a IR called "abstract
bytecode", which are actually higher-order functions and will be executed later
on a abstract abstract machine for that IR. \citet{damian1999partial} provides a
formal treatment to abstract compilation and Shiver's CFA, as well as proofs to
establish correctness of the certified specialized analyzer.
\citet{amtoft1999partial} applied partial evaluation for constraint-based
control flow analysis. Splitting an analysis into multiple stages is also
studied for analysis other than for control-flow, though the formulation may
very different. For example, \citet{DBLP:conf/cgo/HardekopfL11} apply staging to
flow-sensitive pointer analysis. The first stage is to analyze the program code
to obtain a sparse representation, and then the second stage conducts the
flow-sensitive analysis based on the first one.
The abstract domains can be also specialized w.r.t. the static program
structure, e.g., decomposing polyhedras \cite{DBLP:conf/popl/SinghPV17,
Singh:2017:PCD:3177123.3158143} to smaller ones depending on the variables
involved such that the abstract transformers can compute using less time and
space.

\paragraph{Abstract Interpreters} Abstract interpretation was proposed as a
semantic-based approach to construct sound static analysis by approximation
\cite{DBLP:conf/popl/CousotC77}. As semantic artifacts, the Abstracting Abstract
Machines (AAM) \cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10} approach
shows that abstract interpreters can be derived systematically from concrete
semantic artifacts. The AAM approach is closely related to control-flow analysis
for higher-order languages \cite{Midtgaard:2012:CAF:2187671.2187672}.
Using monads to construct abstract interpreters is explored by
\citet{Sergey:2013:MAI:2491956.2491979} and
\citet{DBLP:journals/pacmpl/DaraisLNH17, Darais:2015:GTM:2814270.2814308}.
Similar to definitional abstract interpreters, \citet{Wei:2018:RAA:3243631.3236800}
reconstruct big-step abstract interpreters with delimited continuations;
\citet{Keidel:2018:CSP:3243631.3236767} present a modular arrow-based abstract
interpreter that makes proof of soundness can be constructed compositionally.
\citet{DBLP:conf/cc/CousotC02} proposed an abstract interpretation framework for
modular analysis.

### 2019-04-05 paper/90_conc.tex
paper

In this paper, we first present the implementaion of first Futamura projection
on abstract interpreters. To realize it, we design a generic interpreter that
abstracts over both binding-time and value domains; by instantiating it
differently, the staged abstract interpreter can be readily derived. We also
conduct case studies and empirical evaluation, demonstrating a wide
applicability of our approach and evaluating the performance improved by staging.

>>>>

In this paper, we first present the implementaion of first Futamura projection
on abstract interpreters. To realize it, we design a generic interpreter that
abstracts over both binding-time and value domains; by instantiating it
differently, the staged abstract interpreter can be readily derived.
The meta-programming approach brings new insight and implementation techniques
to optimize abstract interpretation based static analysis. The optimization can
be realized in a compeltely orthogonal way to the abstract semantics and actual
implementation code. We empirical evaluate the performance improved by our approach.
